### 2015-03-28 Kokkos_PG_MachineModel.tex
Added Kokkos model image and some new descriptions of memory spaces/execution spaces.

\chapter{Machine model}\label{C:Model}

@@

\section{Motivation}\label{S:Model:Motivation}

Kokkos is a programming model which happens to have a C++ implementation.
As you are first learning Kokkos,
you may find yourself focusing on details of the implementation,
especially if you have less experience with C++ syntax.
However, as the syntax becomes more familiar to you,
you will find it more and more helpful to understand the underlying abstractions.

Kokkos assumes an \emph{abstract machine model},
which represents a shared-memory parallel computer architecture.
The model may represent either a stand-alone computer,
like a laptop or workstation,
or a single ``parallel process'' in a distributed-memory parallel programming model such as MPI (the Message-Passing Interface).
In general, we write \emph{node} to describe the hardware that Kokkos governs.
Kokkos' machine model lets programmers reason about very complicated heterogenous node architectures.
It also lets them start with simple, even sequential architectures,
and gradually port code from there to perform well on more complicated computer architectures.
The abstract machine model might look complicated,
but it reflects computers that companies are actually designing and building.
Many chapters in this Guide show how to use the more complex aspects of the model.
For example, see Chapter \ref {C:Hierarchical},
which shows how to use Kokkos to expose multiple levels of parallelism.

The point of this chapter is that \emph{respecting the abstract machine model ensures performance portability}.
The phrase \emph{performance portability} means that the same code performs reasonably well on many different computer architectures,
with no or minimal\footnote{For example, users might change a single C++ typedef or a single configure-time option.} changes to code.
Kokkos depends on this abstract machine model for performance portability.
\emph{Implementations} of Kokkos may depend on other things for performance portability.
For example, efficiency of \lstinline!parallel_for! in the C++ implementation relies in part on compilers' ability to inline code.
One could imagine a different Kokkos implementation based on source-to-source translation,
that inlines parallel loop bodies into \lstinline!parallel_for! before they reach the compiler.
Kokkos' OpenMP and Pthreads back-ends may rely on the Portable Hardware Locality (hwloc) library
to control assignment of software threads to hardware cores or hyperthreads.
However, Kokkos as an abstract programming model does not depend on these things.

\section{Memory spaces and execution spaces}\label{S:Model:Spaces}

A \emph{memory space} contains data.
Users access those data as k-dimensional typed arrays, where k may be any nonnegative integer.
(Zero-dimensional arrays are single values.  See Chapter \ref{C:View} for details.)
% In the C++ implementation of the Kokkos programming model,
% these memory addresses are pointers,
% which point to contiguous address ranges,
% each of which represents an allocation of data.
The memory space controls dynamic allocation and deallocation of those arrays,
if it allows them at all.

An \emph{execution space} runs Kokkos' parallel operations,
like parallel for, reduce, and scan.  (See Chapter \ref{C:Dispatch}.)
It may access data in some available memory spaces.
Whether an execution space may access data in a memory space
depends on \emph{both} the execution space and the memory space.

\subsection{Types and instances of spaces}\label{SS:Model:Spaces:Types}

Kokkos distinguishes between \emph{types} of spaces, and \emph{instances} of spaces.
For example, a node might have multiple NVIDIA GPUs.
All of these GPUs' memory spaces have the type \lstinline!Kokkos::CudaSpace!,
but each space is a different instance of \lstinline!Kokkos::CudaSpace!.
The common case of using Kokkos presumes a single instance of a given execution or memory space.
In that case, one uses the type of the space as a synonym for the default (single) instance of that space.
When we speak of a space by its type,
we mean the default single instance of that space,
unless we say otherwise.

\subsection{Analogy with operating system processes}\label{SS:Model:Spaces:Analogies}

Here is an analogy: Many operating systems have ``processes.''
A process is like an execution space, in that code may execute in one or more concurrent threads.
Those threads share access to a common stack and heap, which comprise the process' ``memory space.''
Two different processes may not access each other's memory spaces by default.
However, some operating systems have a ``shared memory'' mechanism.\footnote{See e.g., the POSIX \lstinline!shm_open! and related functions.}
This lets different processes share access to a specific region of memory,
which in Kokkos terms is like a ``memory space,'' separate from each process' normal memory space.
Some operating systems also let you open a file,
and then ``memory map'' that file into a contiguous memory address range.\footnote{See e.g., the POSIX \lstinline!mmap! function.}
This makes the file behave like yet another memory space.
Memory-mapped files may have very different performance characteristics than the usual memory,
even though programmers access both ``memory spaces'' through the same interface.
Whether or not a process may access a file depends on both the process and the file.

% Kokkos' abstract machine model may contain $N$ types of execution spaces
% and $M$ types of memory spaces.
% It is possible that $N \neq M$.

>>>>

\chapter{Machine model}
\label{C:Model}

@@

\section{Motivation}\label{S:Model:Motivations}

Kokkos is comprised of two orthogonal aspects. The first of these is 
an underlying {\em abstract machine model} which describes fundamental
concepts required for the development of future portable and performant
high performance computing applications; the second is a concrete instantiation
of the programming this model written in C++, which allows programmers
to write to the concept machine model. It is important to treat these
two aspects to Kokkos as distinct entities because the underlying model
being used by Kokkos could, in the future, be instantiated in additional
languages beyond C++ yet the algorithmic specification would remain
valid.

\subsection{Kokkos Abstract Machine Model}

\begin{figure}
\includegraphics[width=2in]{figures/kokkos-node.pdf}
\caption{Conceptual Model of a Future High Performance Computing Node}
\label{fig:kokkosnode}
\end{figure}

Kokkos assumes an {\em abstract machine model} for the design of
future shared-memory computing architectures. The model (shown in
Figure~\ref{fig:kokkosnode}) assumes that there may be multiple
execution units in a compute node. In the Figure we have elected to
show two different types of compute units -- one which represents
multiple latency-optimized cores, similar to contemporary processor
cores, and a second source of compute in the form of an off die
accelerator. Of note is that the processor and accelerator each 
have distinct memories, each with unique performance properties,
that are accessible across the node ({\em i.e.} the memory is
reachable or {\em shared} by all execution units. The specific
layout shown in Figure~\ref{fig:kokkosnode} is an instantiation
of the Kokkos abstract machine model used to describe the potential
for multiple types of compute engines and memories within a single
node. In future systems there may be a range of execution engines
which are used in the node ranging from a single type of core
as in many/multi-core processors found today through to an range
of execution units where many-core processors may be joined to
numerous types of accelerator cores. In order to ensure portability
to the potential range of nodes an abstraction of the
compute engines and available memories are required.

\subsection{Kokkos Execution Spaces}

Kokkos uses the term {\em execution spaces} to describe a logical
grouping of computation units which share an indentical set of
performance properties. An execution space provides a set of 
parallel execution resources which can be utilized by the
programmer using several types of fundamental parallel
operation. For a list of the operations available see
Chapter~\ref{C:Dispatch}.

\subsubsection{Execution Space Instances}

An {\em instance} of an execution space is a specific instantiation
of an execution space to which a programmer can target parallel
work. By means of example, an execution space might be used
to describe a multi-core processor. In this example, the
execution space contains several homogeneous cores which share
some logical grouping. In a program written to the Kokkos model,
an instance of this execution space would be made available
on which parallel kernels could be executed. As a second example,
if we were to add a GPU to the multi-core processor a second
execution space type is available in the system, the 
application programmer would then have two 
execution space instances available to select from. The
important consideration here is that the method of
compiling code for different execution spaces and the
dispatch of kernels to instances is abstracted by the Kokkos
framework allowed application programmers to be free from 
written algorithms in hardware specific languages or
requiring the programmer to understand the restrictions
required for parallel execution in the execution space.

\subsection{Kokkos Memory Spaces}

The multiple types of memory which will become available in future
computing nodes are abstracted by Kokkos through {\em memory
spaces}. Each memory space provides a finite storage capacity
at which data structures can be allocated and accessed. In
order to provide portable allocations and performant access,
Kokkos requires data to be allocated as a $k$-dimensioned array,
where $k$ must be greater than equal to zero (a zero array being
a scalar). Each dimension of the array can be either fixed in
size at compile time or dynamically sized with size resolution
performed during execution.

\subsubsection{Instances of Kokkos Memory Spaces}

In much the same way execution spaces have specific instaniations through
the availablilty of an {\em instance} so do memory spaces. An instance
of a memory space provides a concrete method for the
application programmer to request data storage allocatons. Returning
to the examples provided for execution spaces, the multi-core
processor may have multiple memory spaces available including
on-package memory, slower DRAM and additional a set of non-volatile
memories. The GPU may also provide an additional memory space
through its local on-package memory. The programmer is free
to decide where each data structure may be allocated by requesting
these from the specific instance associated with that memory space.
Kokkos provides the appropriate abstraction of the allocation
routines and any associated data management operations including
releasing the memory and returning it for future use, as well as
copy operations.
